Inferring probabilistic stellar rotation periods using Gaussian processes

### 2017-02-13 documents/ms.tex
minor edits

Not all \kepler\ light curves show evidence of stellar rotation.
In some cases a star may have few or no active regions, be rotating
pole-on, or be rotating so slowly that the \kepler\ data detrending pipeline
removes any signal.
In other cases there may be another source of variability present in the light
curve, generating a false period detection.
These sources may be physical: \eg\ binary star interactions, intra-pixel
contamination from other astrophysical objects, pulsating variable stars,
asteroseismic oscillations in giants and even stellar activity cycles.
For some of these there is little we can do save attempt to eliminate these
astrophysical false positives via alternative methods, \eg\ apply colour cuts
to avoid giant contamination.
For some however, like the variable stars, they may have distinctive
hyperparameters that identify them, for example long coherence timescales.
Testing this is beyond the scope of this paper but may be an interesting
follow-up study.
As well as astrophysical contamination, there may also be instrumental sources
of contaminating variability: \eg\ temperature variations or pointing shifts
of the \kepler\ spacecraft.
These are unlikely to be periodic and, again, may produce unusual combinations
of hyperparameters.
This is something that we hope to test in future.
In addition to this, there are several other aspects of the GP method we are
continuing to develop, listed as follows:

@@

\item{To perform model selection with different kernel functions. Again, these
    tests have not been performed thus far due to the cost of calculating the

@@

        kernel function, $P$ but some others may warrant physically motivated

@@

        For large values of $l$ ($l\gg P$), $\Gamma$ can be related to the
        typical number of stationary/non-stationary points of inflection, per

@@

        In typical light curves, observing more than say three or four

@@

        (as opposed to say $20$; of course $2$ turning points is the minimum

@@

Another huge advantage of the GP method is, because it is a {\it generative}
model of the data, the rotation period signal can be modelled at the same time
as systematic noise features.

@@

parameters of interest will insure that the periodic signal is preserved.}

@@

We attempted to recover the rotation periods of \naigrain\ simulated
\kepler-like light curves for solid-body rotators \citep{Aigrain2015} using
three different methods: a new Gaussian process method, a Lomb-Scargle
periodogram method and an autocorrelation function method.
We demonstrate that the GP method produces the most accurate rotation periods
of the three techniques, providing a large improvement over the LS periodogram
method and a moderate one over the ACF method.
We also find that the standard version of the ACF method, most commonly
implemented in the literature, often slightly under-predicts the rotation
period due to a subtle feature of its peak detection algorithm.
In addition, we compared the rotation periods of \kepler\ objects of
interest measured using the GP method to those measured previously by
\citet{Mcquillan2013}.
The good agreement between the two sets of results demonstrates that the GP
method works well on real \kepler\ data.
Unlike the ACF and LS periodogram methods, the GP method provides posterior
PDF samples which can be used to estimate rotation periods uncertainties.
Although an improvement on competing methods, these uncertainties are still
underestimated in many cases because the GP model is an approximate, effective
model: we find that only a quarter of the reported uncertainties are accurate.
% in many cases the posterior PDFs of the parameters are
% multi-modal.
% If it were possible to run an MCMC sampler for an infinite number of steps,
% the entire posterior PDF would eventually be sampled, and a more accurate
% uncertainty would result.
% Unfortunately, due to realistic computational resources and the limitations of
% MCMC sampling techniques for multi-modal posteriors, we are not always able to
% map out the entire posterior PDFs.
% \tdmcomment{I don't think this is the problem;
% I think the fits that we judge to have converged have actually converged.}
In addition, although the GP model used here is clearly a good one, it is
still only an effective model, not an accurate physical model.
It can only capture the posterior PDF of the periodic component of the
covariance matrix, not the actual rotation period of a physical star.

>>>>

Not all \kepler\ light curves show evidence of stellar rotation.
In some cases a star may have few or no active regions, be rotating
pole-on, or be rotating so slowly that the \kepler\ data detrending pipeline
removes any signal.
In other cases there may be another source of variability present in the light
curve, generating a false period detection.
These sources may be physical: \eg\ binary star interactions, intra-pixel
contamination from other astrophysical objects, pulsating variable stars,
asteroseismic oscillations in giants and even stellar activity cycles.
Identifying many of these astrophysical false positives 
falls outside the scope of this GP method 
(\eg\ applying colour cuts to avoid giant contamination);
however, for some, such as variable stars, they may have distinctive
hyperparameters (\eg\ long coherence timescales) that identify them.
Testing this may be an interesting follow-up study.
As well as astrophysical contamination, instrumental sources may contribute
to contaminating variability, \eg\ temperature variations or pointing shifts
of the \kepler\ spacecraft.
These are unlikely to be periodic and, again, may produce unusual combinations
of hyperparameters.
We also hope to test this in the future.

In addition, we are continuing to develop several other aspects of this GP method:

@@

\item{To perform model selection with different kernel functions. 
    We have not performed these tests yet due to the cost of calculating the

@@

        kernel function ($P$), but others may warrant physically motivated

@@

        For large values of $l$ ($l\gg P$), $\Gamma$ relates to the
        typical number of stationary/non-stationary points of inflection per

@@

        In typical light curves, observing more than perhaps three or four

@@

        (of course, $2$ turning points is the minimum

@@

Because it is a {\it generative} model of the data,
the GP method models the rotation period at the same time as systematic
noise features.

@@

parameters of interest will insure that the periodic signal is preserved. 
\tdmcomment{Is this bullet point obselete, given that we have already successfully
marginalized over \kepler\ noise to recover periods? Or is there
something else we wanted to say here?}}

@@

We have attempted to recover the rotation periods of \naigrain\ simulated
\kepler-like light curves for solid-body rotators \citep{Aigrain2015} using
three different methods: a new Gaussian process method, 
an authocorrelation function method, and a Lomb-Scargle periodogram method.
We demonstrate that the GP method produces the most accurate rotation periods
of the three techniques, providing a large improvement over the LS periodogram
method and a moderate one over the ACF method.
We also find that the standard version of the ACF method, most commonly
implemented in the literature, often slightly under-predicts the rotation
period due to a subtle feature of its peak detection algorithm.
In addition, we measure the rotation periods of \nkoimcq\ \kepler\ objects of
interest using the GP method, and find that these results compare well
to those measured previously by \citet{Mcquillan2013}.
The good agreement between these two sets of results demonstrates that the GP
method works well on real \kepler\ data.

Unlike the ACF and L-S periodogram methods, the GP method provides posterior
PDF samples which can be used to estimate rotation periods uncertainties.
Although an improvement on competing methods, these uncertainties are still
underestimated in many cases because the GP model is an approximate, effective
model. \tdm{Can we make a statement something like ``we believe the method underestimates
uncertainties by about a factor of three'' or something like that?  This
matches with 66\% of the fits being within ``3$\sigma''.
Might be good to say something like this, both in the discussion and here.}
% in many cases the posterior PDFs of the parameters are
% multi-modal.
% If it were possible to run an MCMC sampler for an infinite number of steps,
% the entire posterior PDF would eventually be sampled, and a more accurate
% uncertainty would result.
% Unfortunately, due to realistic computational resources and the limitations of
% MCMC sampling techniques for multi-modal posteriors, we are not always able to
% map out the entire posterior PDFs.
% \tdmcomment{I don't think this is the problem;
% I think the fits that we judge to have converged have actually converged.}
In addition, although the GP model used here is clearly a good one, it is
still only an effective model, not an accurate physical model.
It can only capture the posterior PDF of the periodic component of the
covariance matrix, not the actual rotation period of a physical star.
